### 2012-01-22 paper_draft.tex
some changes in the introduction

Genome-Wide Association Studies (GWAS) are a comparatively easy and cheap way of finding Single Nucleotide Polymorphisms (SNPs) of medical relevance. Such SNPs of interest can be used to find candidate genes for a closer inspection or to predict disease risks or other traits. Genome-Wide Association Studies make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This method does not allow inference of causal differences but merely identifies correlations. The first GWAS was published in 2005 and compared age-related macular degeneration in contrast to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. %(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)
\\
\\
FZ REWRITE 
\\
The comparison of DNA sequences from indivduals in a population reveals sites that are variable. 
These variations are called Single Nucleotide Polymorphisms (SNP). Each different nucleotide 
at a specific site is defined as an allele. FZ: Explain genotype!!
Different methods have been developed to measure/find/access/sequence SNPs in an individual, a
process called genotyping. Microarrays are a cheap and fast way to genotype individuals, in comparison
to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associated SNPs with for example diseass. Genome-Wide Assocation Studies
(GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
enables GWAS to find significant overrepresantations of alleles in patients and associate them wir a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

FZ: The data is scattered and unorganized, making it hard to use this infornomation. While projects like the SNPedia try to keep track of all the files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data.

@@

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their familiy members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).

@@

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
\subsubsection*{Why should a user publish their genetic informationß Pro vs. Con}
With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising. 
\subsubsection*{Privacy and health implications and ethical considerations}

>>>>

The comparison of DNA sequences from individuals in a population reveals sites which are variable. 
If differences between individuals span only over a single nucleotide then these variations are also called Single Nucleotide Polymorphisms (SNP) and different nucleotides 
at a specific site are defined as an allele. For diploid organisms, as us humans, the genotype at a given site consists of two alleles, which are independently inherited from each parent.
Different methods have been developed to read SNPs in an individual, a
process called genotyping. Microarrays are a cheap, fast and widespread used way to genotype individuals, in comparison
to whole genome sequencing. If the genotypes for more then one individual are known, the analysis of
allele frequencies can be used to associated SNPs with diseases and other inheritable traits. Genome-Wide Assocation Studies
(GWAS) make use of statistics for comparing the alleles of patients to the alleles of healthy controls. This
enables GWAS to find alleles which are significantly overrepresented in patients and associate them with a disease.
This method does not allow inference of causal differences but merely identifies correlations. 
The first GWAS was published in 2005 and compared age-related macular degeneration in contrast 
to a healthy control group \cite{Klein2005}. Since the beginning, the number of participants in 
such studies has been rising: Over 1200 GWAS have been performed \cite{Johnson2009} and over 
5000 SNPs have been linked to different diseases and traits in those studies \cite{Hindorff2009}. 
%(http://www.genome.gov/page.cfm?pageid=26525384&clearquery=1#result_table)

@@

This data is scattered and unorganized, making it hard to use this information. While projects like SNPedia try to keep track of all the genotyping files \cite{Cariaso2011}, they do not provide the information necessary to perform GWAS, as the phenotypic information is most often not attached to the genetic information. Projects that attach the phenotype to the genetic information, like the \textit{Personal Genome Project}, still don't allow for an easy re-use of the data, as they lack an advanced programming interface (API) or other methods by which researchers could download the data and not every customer of DTC genetic testing can participate in it.

@@

Those participants who are not willing to share data with their DTC company are likely to agree more strongly with the following motivations for not sharing their data when compared to those who would share their data with their DTC company under any circumstances: because they fear discrimination (mean difference = 1.52, SE = 0.322, p < 0.001), because they feel it is a breach of their privacy (mean difference = 1.871, SE = 0.324, p < 0.001), because they fear consequences for their family members (mean difference = 1.146, SE = 0.32, p = 0.001) and because they fear personalized advertising (mean difference =  1.112, SE = 0.357, p = 0.006).

@@

Another risk regarding data quality that should be kept in mind is a possible bias in data availability on openSNP, as we cannot rule out the possibility that only an unrepresentative subset of people buy DTC genetic testing and an even smaller subset is willing to publish the results, along with information of medical relevance.
\subsubsection*{Why should a user publish their genetic information? Pro vs. Con}
With openSNP, we have built a platform that can be used by customers of DTC genetic testing to easily share their genetic and phenotypic data with a wide audience, as well as by scientists and interested citizens who are looking for datasets to be used in GWAS. Customers of DTC genetic testing also benefit from an easy access to primary literature on SNPs and genetic variations they carry. While we don't have collected enough data to perform full scale GWAS yet, this might be possible in the future, as user numbers are rising. 
\subsubsection*{Privacy and health implications and ethical considerations}
